Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2010-08-10 Text/migration.tex
First pass over Dick's draft. Need more bibliography work and text tweaking.
	new file:   .gitignore
	modified:   Text/migration.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@68 3b7f7548-5c66-0410-83ab-3997167635fc

Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that is massive enough to experience gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.13 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially only marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases, there are times when the planet's radial motion is slow and random. At other times, when the planet interacts with a strong spiral mode near the mode's corotation radius, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\bf See note in Latex file.}
%Scott and Aaron: Verify that everything I say here is correct.
The CHYMERA (Computational HYdrodynamics with MultiplE Radiation Algorithms) code (Boley et al. 2007b) is a second-order, explicit, Eulerian scheme on a 3D cylindrical grid. The code uses a realistic equation of state for H$_2$ (Boley et al. 2007a) and integrates an energy equation that includes $PdV$ work, net heating or cooling due to radiative flux divergence, and heating by artificial bulk viscosity. Calculations are done on a uniform cylindrical grid with reflection symmetry about the disk midplane and a resolution ($\varpi$,$\phi$,z) = (512,512,64). The $z$-axis is the rotation axis of the disk. The large number of azimuthal zones is necessary to resolve the planet's Hill sphere and the planet's wake. These simulations utilize the radiative cooling scheme developed and tested in Boley et al. (2007b), where the optically thick monochromatic flux in the $\varpi$ and $\phi$-directions is computed by flux-limited radiative diffusion and where the radiative transport of energy in the $z$-direction is solved using a one-ray discrete ordinate method in both optically thin and thick regions. Although the central star remains fixed at the grid center, we account for acceleration of the reference frame by the planet and by the disk via indirect potentials, as in Michael \& Durisen (2010), and, overall, the planet integration is done using methods similar to those in Nelson et al. (2000). The Rosseland mean and Planck mean opacities and molecular weights in our simulations are the same as those in Boley et al. (2006, 2007b), except that we correct the D'Alessio et al. (2001) gas mean molecular weight to $\mu$ = 2.39.

@@

The model disk, based on an equlibrium model in Pickett et al. (2003), orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.13 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 250 yr. The disk is initially located between radial grid zones ??? 30 ??? and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable (see Durisen et al. 2007) minimum value of ??? at radial grid zone ??? (??? AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. There is an inner cylindrical hole in the grid with a radius of 13 radial zones. To seed nonaxisymmetry, the density is given an initial ??? \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the fiducial run, goes from time $t = 0$ to ??? 20 ??? ORP without a planet. In the second simulation, which we call the $t = 0$ planet run, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the $t = 10$ planet run, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10$ ORP. These three simulations are part of a much larger suite, to be described elsewhere, in which the planet mass and starting locations are varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by finding at the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, over each interval that the planet's $\phi$ changes by 2$\pi$ (one ``orbit'') and then using $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These orbital elements are only intended to be indicative; there are strong orbit to orbit variations in these $a$ and $e$.

@@

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities (Li and Lovelace articles abound 2000, 2001). Meschiari \& Laughlin (2008) recently demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs.

%Scott and Aaron: Can you go over this part carefully and fill in some calculation for the migration rate?
Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times1o^{-3}$ AU yr$^{-1}$. Although there are no formulae that are fully applicable to our radiatively cooled disk, using equation ??? from ??? and parameters from the simulated disk, we estimate ??? AU yr$^{-1}$ as an analytic prediction. While the interaction with a GI-active disk did not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in the GI-active disk is highly variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In the case of the $t = 10$ planet run, the overall radial migration rate is similar but highly erratic. The important departures from laminar disk theory are that there are times when migration stalls and that interactions with discrete low-order modes seem to be important. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CR of low-order GI modes (for $t = 10$ type runs).

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research was conducted with the support of NASA grants from the Origins of Solar Systems  Program (NNG05GN11G and NNX08AK36G). This material is based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and CNS-0521433. Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation.

>>>>

Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that is massive enough to experience {\bf any mass object will experience GIs rewrite} gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially marginally stable. In one simulation, the planet is inserted prior to the initial eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases, there are times when the planet's radial motion is slow and random. At other times, when the planet interacts with a strong spiral mode near the mode's corotation radius, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. {\bf See note in Latex file.}
%The assertion in the preceding sentence needs to be verified by analyzing both runs in more detail -- Am(t), periodograms, images zoomed in on the planet to check for arms. The migration is observed; it is the relationship to mode interactions that needs to be checked.
Planet orbit eccentricity tends to fluctuate rapidly and ranges between about 0.02 to 0.1.
\end{abstract}

@@

{\bf See note in Latex file.}
%Scott and Aaron: Verify that everything I say here is correct.
The CHYMERA (Computational HYdrodynamics with MultiplE Radiation Algorithms) code (Boley et al. 2007b) is a second-order, explicit, Eulerian scheme on a 3D cylindrical grid. The code uses a realistic equation of state for H$_2$ (Boley et al. 2007a) and integrates an energy equation that includes $PdV$ work, net heating or cooling due to radiative flux divergence, and heating by artificial bulk viscosity. Calculations are done on a uniform cylindrical grid with reflection symmetry about the disk midplane and a resolution {\bf starting resolution is different} ($\varpi$,$\phi$,z) = (512,512,64). The $z$-axis is the rotation axis of the disk. The large number of azimuthal zones is necessary to resolve the planet's Hill sphere and the planet's wake. These simulations utilize the radiative cooling scheme developed and tested in Boley et al. (2007b), where the optically thick monochromatic flux in the $\varpi$ and $\phi$-directions is computed by flux-limited radiative diffusion and where the radiative transport of energy in the $z$-direction is solved using a one-ray discrete ordinate method in both optically thin and thick regions. Although the central star remains fixed at the grid center, we account for acceleration of the reference frame by the planet and by the disk via indirect potentials, as in Michael \& Durisen (2010), and, overall, the planet integration is done using methods similar to those in {\bf integration is done with a verlet integrator as in hut95, yoshida90, channell90, tuckerman92. indirect terms are treated as in nelson} Nelson et al. (2000). The Rosseland mean and Planck mean opacities and molecular weights in our simulations are the same as those in Boley et al. (2006, 2007b), except that we correct the D'Alessio et al. (2001) gas mean molecular weight to $\mu$ = 2.39.

@@

The model disk, based on an equlibrium model in Pickett et al. (2003), orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 250 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable (see Durisen et al. 2007) minimum value of ??? at radial grid zone ??? (??? AU). The computational grid is extended radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. There is an inner cylindrical hole in the grid with a radius of 12 radial zones {\bf brief description of hole needed}. To seed nonaxisymmetry, the density is given an initial 0.01 \% random cell-to-cell perturbation.

This Letter presents three simulations. The first, which we call the fiducial run, goes from time $t = 0$ to ??? 20 ??? ORP without a planet. In the second simulation, which we call the T0 planet run, a one Jupiter-mass ($M_J$) planet with a roughly circular midplane orbit is included in the initial $t = 0$ ORP equilibrium disk at 25 AU. In the third simulation, which we call the T10 planet run, a 1 $M_J$ planet with a similar orbit is inserted into the fiducial run at $t = 10$ ORP. These three simulations are part of a much larger suite, to be described elsewhere, in which the planet mass {\bf and starting locations} are varied. We estimate crude semi-major axes $a$ and eccentricities $e$ from the planet's $\varpi(t)$ by finding at the maximum and minimum radii, $\varpi_{\rm max}$ and $\varpi_{\rm min}$, over each interval that the planet's $\phi$ changes by 2$\pi$ (one ``orbit'') and then using $a = (\varpi_{\rm max}+\varpi_{\rm min})/2$ and $e = (\varpi_{\rm max}-\varpi_{\rm min})/(\varpi_{\rm max}+\varpi_{\rm min})$. These orbital elements are only intended to be indicative; there are strong orbit to orbit variations in these $a$ and $e$.

@@

Even in non-self-gravitating disks, the presence of discrete radial structure is known to cause instabilities (Li and Lovelace articles abound 2000, 2001). Meschiari \& Laughlin (2008) recently demonstrated that a gap, presumably opened by an embedded planet, can cause global modes to become unstable in self-gravitating disks that are otherwise stable against the development of GIs. Here we are dealing with a planet that is not massive enough to open a gap, but its gravitational interaction with the disk also has a strong effect on the growth and onset of GIs. {\bf say more about gap opening criteria and whether they apply here}

%Scott and Aaron: Can you go over this part carefully and fill in some calculation for the migration rate?
Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no formulae that are fully applicable to our radiatively cooled disk, using equation ??? from ??? and parameters from the simulated disk, we estimate ??? AU yr$^{-1}$ as an analytic prediction. While the interaction with a GI-active disk did not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in the GI-active disk is highly variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In the case of the $t = 10$ planet run, the overall radial migration rate is similar but highly erratic. The important departures from laminar disk theory are that there are times when migration stalls and that interactions with discrete low-order modes seem to be important. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CR of low-order GI modes (for $t = 10$ type runs).

@@

The authors would like to thank F.C. Adams and J.C.B. Papaloizou for useful encouragement in this project. The research was conducted with the support of NASA grants from the Origins of Solar Systems  Program (NNG05GN11G and NNX08AK36G). This material is based upon work supported by the National Science Foundation under Grants No. ACI-0338618l, OCI-0451237, OCI-0535258, OCI-0504075, and CNS-0521433. Any opinions, findings and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views of the National Science Foundation. {\bf need TG acks}
